Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-19 grove-paper.tex
Revise Section 2.3

Having described single-user actions,
we turn our attention to how Grove handles multiple users.
This section generalizes to any number of users,
but for simplicity we consider only two: Alice and Bob.
Alice and Bob each maintain their own editor instance and perform
actions relative to its state (using their own cursors, which we discuss in \autoref{sub:Cursors}).
We start each user where we left off in the previous section with the state shown in \autoref{fig:Move:b}.

@@

To start, Alice and Bob both have the graph in \autoref{fig:Move:b}, which
represents the code~\texttt{\hole{} + (\hole{} * y)}.

Alice adds~\vDifferentPartsAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vSimpleY{} to \vDifferentPartsBob{}.
Before transmitting their edits to each other,
Alice and Bob have the graphs in \autoref{fig:DifferentParts:a}
and \autoref{fig:DifferentParts:b}, respectively.
(Note that the transition edge from \autoref{fig:Move:b} to
\autoref{fig:DifferentParts:b} represents
multiple edges, i.e., deleting~\eSimpleY{} and adding~\eDifferentPartsBob{} along with its child~\vDifferentPartsBob{}.
We thus mark it with a star.)\todo{star size}\todo{"sync"}

@@

Stated more formally, our edit model obeys the commutativity shown in \autoref{fig:Commutativity}.
For any sequences of edits from Alice and Bob,
applying them in either order produces the same results.

@@

After Alice and Bob share the edits they made in the previous subsection,
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{}, which results in the graph in \autoref{fig:NestedParts:a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the right child of~\vWrapPlus{} to its left child
by deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}, which results in the graph in \autoref{fig:NestedParts:b}.
Since edits are based on edge identities,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:NestedParts:c}.

A consequence of this design is that we allow users to modify
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of~\eMoveTimes{} before making her
changes and before receiving Bob's addition of~\eNestedPartsBob{}.
Her edits would then temporarily be within the code deleted by Bob.
TODO: \todo{TODO}She can restore that code using a restoration action add a fresh edge to it.

TODO:figure

@@

Next we address what happens when edits conflict.
For example, Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in Alice and Bob having the graphs in \autoref{fig:MultiChild:a} and \autoref{fig:MultiChild:b}, respectively.
These graphs have two different edges for the~\texttt{R} position of~\vWrapPlus{}.
\autoref{fig:MultiChild:a} has~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and \autoref{fig:MultiChild:b} has~\eMultiChildBob{}, which points to~\vMultiChildBob{}.
We call this a \emph{multi-child conflict}.

We merge these edits by including \emph{both} edges as edges going out of the \texttt{R} position
of \vWrapPlus{}.
This results in \autoref{fig:MultiChild:c}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index.
In the user interface, we flag this conflict with the notation~\texttt{\{ x | y \}}.

This conflict must eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \eMultiChildBob{} and then change~\vMultiChildAlice{}
to be~\texttt{x * y}.
(Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
  may be possible to develop evaluation models that allow these sorts of conflicts,
  but that is beyond the scope of this paper.})

@@

TODO: Multiple adds cause conflicts (Not usually presented in the UI)

Our tree-based model is similar in that it presents alternate versions
of the code, and it is up to the user to replace those alternates with the desired merged result.
Our model is different in that merge conflicts in the line-based model
prevent some semantic editor services from running
that could continue to run in the tree-based model because
code inside and outside the conflict is still valid code.
It is just the point where they meet the poses a potential problem for editor services.
Another place our model is different
is that changes are explicit rather than being inferred by a \texttt{diff} algorithm.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the grammatical structure of the code.

@@

In addition to conflicts due to multiple edges coming \emph{out of} the same child position of a particular vertex,
conflicts can manifest as multiple edges going \emph{into} a particular vertex.
We call these \emph{multi-parent conflicts}.

An example of this is shown in \autoref{fig:MultiParent}.
\autoref{fig:MultiParent:a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w * \hole{}) + \hole{}}.
Alice moves~\vNestedPartsAlice{} to the~\texttt{R} position of~\vSimpleTimes{},
while Bob moves it to the~\texttt{R} position of~\vWrapPlus{}.
This results in the graphs in \autoref{fig:MultiParent:b} and \autoref{fig:MultiParent:c},
which represent respectively the expression~\texttt{(\hole{} * w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} * \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{}.
As explained in \autoref{sec:Formalism}, deletes are idempotent so Alice and Bob both deleting~\eNestedPartsAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:MultiParent:d}, in which~\vNestedPartsAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

@@

For example, consider the situation in \autoref{fig:Cycle}.
\autoref{fig:Cycle:a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole{} * \hole{}) * (\hole{} + \hole{})) + \hole}.
Alice moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath.
This results in \autoref{fig:Cycle:b},
which represents the expression~\texttt{(\hole{} * \hole{}) + ((\hole{} + \hole{}) * \hole{})}.
On the other hand, Bob does the same
but puts~\vMultiCycleTimes{} under~\vMultiCyclePlus{}.
This results in \autoref{fig:Cycle:c},
which represents the expression~\texttt{(\hole{} * \hole{}) + ((\hole{} * \hole{}) + \hole{})}.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Cycle:d}, which has a cycle
between~\vMultiCycleTimes{} and~\vMultiCyclePlus{}.
This situation is akin to a merge conflict in traditional version control systems.
Resolving it requires user input, but is as simple as
either deleting~\eMultiCycleBobPlus{} and~\eMultiCycleBobTimes{}~(thus favoring Alice's version)
or deleting~\eMultiCycleAliceTimes{} and~\eMultiCycleAlicePlus{}~(thus favoring Bob's version).
Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Cycle:d} would be displayed.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

\subsubsection{Disconnection}
\label{sub:Disconnection}

@@

@@

Finally, when multi-user edits are merged,
parts of the graph can become disconnected
even though they are connected in each user's
copy of the graph before the merge.

For example, consider the situation in \autoref{fig:Disconnection}.
\autoref{fig:Disconnect:a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(\hole{} * \hole{}) + (\hole{}  * \hole{})}.
Alice moves~\vMultiCycleTimes{} under~\vSimpleTimes{}.
This results in \autoref{fig:Disconnect:b},
which represents the expression~\texttt{(\hole{} * (\hole{} * \hole{})) + \hole}.
On the other hand, Bob puts~\vSimpleTimes{} under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnect:c},
which represents the expression~\texttt{((\hole{} * \hole{}) * \hole{}) + \hole{}}.

When these edits are merged, the result is the graph in \autoref{fig:Disconnect:d}.
Since Alice deleted~\eNestedPartsBob{}
and Bob deleted~\eMultiCycleAliceTimes{},
when these are merged~\vMultiCycleTimes{} and~\vSimpleTimes{}
are disconnected from the root.

@@

Fortunately, we can detect this situation and warn the user
since the disconnection appears after merging two graphs
that did not have disconnected vertexes.

@@

\subsection{Commutativity}

@@

>>>>

Having described single-user actions,
we turn our attention to how Grove handles multiple users.
This section generalizes to any number of users,
but for simplicity we consider only two: Alice and Bob.
Alice and Bob each maintain their own editor instances and perform
actions relative to its state (using their own cursors, which we discuss in \autoref{sub:Cursors}).

@@

We start each user where we left off in the previous section with the state shown in \autoref{fig:Move:b}.
Alice adds~\vDifferentPartsAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vSimpleY{} to \vDifferentPartsBob{}.
Before transmitting their edits to each other,
Alice and Bob have the graphs in \autoref{fig:DifferentParts:a}
and \autoref{fig:DifferentParts:b}, respectively.
(Note that the transition edge from \autoref{fig:Move:b} to
\autoref{fig:DifferentParts:b} represents
multiple edges, i.e., deleting~\eSimpleY{} and adding~\eDifferentPartsBob{} along with its child~\vDifferentPartsBob{}.
We thus mark it with a star.)

@@

Stated more formally, our edit model obeys the commutativity shown in \autoref{fig:Commutativity}.
For any sequences of edits from Alice and Bob,
applying them in any order always produces the same results.
(We show this formally in \autoref{sec:Commutativity}.)

@@

After Alice and Bob share their edits to get the graph in \autoref{fig:DifferentParts:c},
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{}, which results in the graph in \autoref{fig:NestedParts:a}.
Meanwhile, Bob moves~\vSimpleTimes{} and its children from the~\texttt{R} child of~\vWrapPlus{} to its~\texttt{L} child
by deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}, which results in the graph in \autoref{fig:NestedParts:b}.
Since edits are based on edge identifiers and not a location in the tree,
both Alice's and Bob's edits can be applied to the other's graph,
which results in Alice and Bob both having the graph in \autoref{fig:NestedParts:c}.

A consequence of this commutativity is that users can modify
deleted code.  For example,
Alice might receive Bob's deletion of~\eMoveTimes{} before making her
changes and before receiving Bob's addition of~\eNestedPartsBob{}.
Her edits would then temporarily be within the code deleted by Bob.
TODO: \todo{TODO}She can restore that code using a restoration action add a fresh edge to it.

\subsection{Merge Conflicts}
\label{sub:Merge Conflicts}

While the collaborative edits discussed so far merge cleanly,
they can also generate graphs that do not map cleanly to a syntax tree.
We call these conflicts and consider cases for \emph{multi-child},
\emph{multi-parent}, and \emph{cycle} conflicts.\todo{disconnect}

TODO: as with conflict in line/diff-based version control system (e.g., \texttt{git}, \texttt{mercurial}, \texttt{svn}, etc.),
these require user intervention to resolve.

@@

As an example of a multi-chid conflict, suppose Alice and Bob both have the graph in \autoref{fig:NestedParts-c}.
Alice then adds~\vMultiChildAlice{} as the~\texttt{R} child of~\vWrapPlus{}
while Bob also adds~\vMultiChildBob{} as the~\texttt{R} child of~\vWrapPlus{}.
These result in Alice and Bob having the graphs in \autoref{fig:MultiChild:a} and \autoref{fig:MultiChild:b}, respectively.
Note that these graphs have two different edges for the~\texttt{R} position of~\vWrapPlus{}.

We merge these edits by including \emph{both} edges as edges going out of the \texttt{R} position
of \vWrapPlus{}.
This results in \autoref{fig:MultiChild:c}, where there is a multi-child conflict
in the~\texttt{R} position of~\vWrapPlus~(i.e., there are multiple edges for the same position in a vertex).
In the user interface, we flag this conflict with the notation~\texttt{\{ x | y \}}.

This conflict can be resolved by the user by deleting
one of the edges in conflict and editing expression pointed to the other edge to be the correctly merged value.
For example, one could delete \eMultiChildBob{} and then change~\vMultiChildAlice{} to be~\texttt{x * y}.
(Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
  may be possible to develop evaluation models that allow these sorts of conflicts,
  but that is beyond the scope of this paper.})

@@

This handling of conflicts is similar traditional merge conflicts in that
it presents alternate versions of the code that the user replaces with the desired merge result.
Our model is different in that merge conflicts in the line-based model
prevent semantic editor services from running
that could continue to run in the graph-based model because
code inside and outside the conflict is still valid code.
It is just the where the multi-child happens that editor services may break down.
\autoref{sec:Related Work} discusses more differences.

@@

While multi-child conflicts are when multiple edges are in the same \emph{child} position of a vertex,
multi-parent conflicts are when there are multiple \emph{parents} of a vertex.
An example of this is in \autoref{fig:MultiParent}.
\autoref{fig:MultiParent:a} shows the graph after Alice and Bob have made some edits,
and Alice and Bob's shared their edits with each other.
Alice then moves~\vNestedPartsAlice{} to the~\texttt{R} position of~\vSimpleTimes{},
while Bob moves it to the~\texttt{R} position of~\vWrapPlus{}.
This results in the graphs in \autoref{fig:MultiParent:b} and \autoref{fig:MultiParent:c}.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{}.
As explained in \autoref{sec:Formalism}, deletes are idempotent so Alice and Bob both deleting~\eNestedPartsAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
When their edits are merged, the result is \autoref{fig:MultiParent:d}, in which~\vNestedPartsAlice{} has two parents pointing to it.

@@

TODO: technically a multi-parent conflict

@@

In addition to multi-child and multi-parent conflicts, merges
can cause cycles in the graph.\footnote{In theory a single user could
generate cycles, but this could be detected immediately and the user action
that caused that cycle rejected.
Thus, we are only concerned with cycles that cannot be immediately detected
and prevented by the editor.}
For example, consider the situation in \autoref{fig:Cycle}.
\autoref{fig:Cycle:a} shows the graph after Alice and Bob have made some edits,
and shared them with each other.
Alice then moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath that.
Bob does the same but puts~\vMultiCycleTimes{} under~\vMultiCyclePlus{}.
This results in \autoref{fig:Cycle:b} and \autoref{fig:Cycle:c}, respectively.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Cycle:d}, which has a cycle
between~\vMultiCycleTimes{} and~\vMultiCyclePlus{}.
As with the other kinds of conflict, resolving it requires user input, but is as simple as
either deleting~\eMultiCycleBobPlus{} and~\eMultiCycleBobTimes{}~(thus favoring Alice's version)
or deleting~\eMultiCycleAliceTimes{} and~\eMultiCycleAlicePlus{}~(thus favoring Bob's version).
Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Cycle:d} would be displayed.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

\paragraph{Disconnected Cycles}

@@

TODO: technically a cycle

TODO: in general detect when there are or were edits to something now deleted

@@

Finally, when multi-user edits are merged,
parts of the graph can become disconnected from the root
even though they are connected to the root in each user's
copy of the graph before the merge.

For example, consider the situation in \autoref{fig:Disconnection-a},
which shows the graph after Alice and Bob have made some edits,
and shared their edits with each other.
Alice then moves~\vMultiCycleTimes{} under~\vSimpleTimes{},
while Bob moves~\vSimpleTimes{} under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnect:b} and \autoref{fig:Disconnect:c}, respectively.

When these edits are merged, the result is \autoref{fig:Disconnect:d}.
Since Alice deleted~\eNestedPartsBob{}
and Bob deleted~\eMultiCycleAliceTimes{},
when these are merged~\vMultiCycleTimes{} and~\vSimpleTimes{}
are disconnected from the root.

@@

Fortunately, we can detect this situation and warn the user
since the disconnection appears after merging two graphs
in which vertexes were not already disconnected.

@@

\subsection{Commutativity}
\label{sub:Commutativity}

@@

Another place our model is different
is that changes are explicit rather than being inferred by a \texttt{diff} algorithm.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the grammatical structure of the code.
